---
title: "Hurricanes in the Atlantic"
subtitle: "An assessment of the factors driving the costs of hurricane damage."
author: "Carmen Hoyt"
date: last-modified
editor_options: 
  chunk_output_type: console
execute:
  warning: false
  messages: false
format:
  html:
      toc: true
      code-fold: true
---

# About

Growing up on the Gulf Coast of Florida, my family and I were no stranger to hurricanes. We have lived in the same house for over 20 years, but recent hurricane seasons[^1] have brought an increase in flooding and severe damage. For example, Hurricane Ian (2022) stands out as having been the worst storm to hit our neighborhood during my lifetime, bringing an unprecedented 7 feet of storm surge into our home. 

[^1]: Hurricane season runs June 1 to November 30, but since 2021, the [National Hurricane Center (NHC) has considered moving the start date to May 15](https://www.accuweather.com/en/hurricane/nhc-considering-change-to-start-date-of-hurricane-season/1168271) to encompass increasing early-season activity.

Climate change is influencing natural disasters across the board, and hurricanes are no exception. While the frequency of storms is not projected to change, they are predicted to become more intense[^2]. Hurricane "intensity" is measured in categories on the Saffir-Simpson scale: 1 through 5 based on wind speed[^3]. Storms assigned a category 3 or higher are considered "major hurricanes" and are expected to bring catastrophic damage (and thus more significant damage costs). This leads to my question:

**As storms increase in intensity, are damage costs also being driven by time?**

[^2]: Read more about how hurricanes are expected to evolve with climate change [here](https://science.nasa.gov/earth/climate-change/a-force-of-nature-hurricanes-in-a-changing-climate/).

[^3]: Read more about the Saffir-Simpson scale [here](https://www.weather.gov/mfl/saffirsimpson#:~:text=The%20Saffir%2DSimpson%20Hurricane%20Wind,loss%20of%20life%20and%20damage.).

# Data

I will be using Kaggle's [North American Hurricanes from 2000](https://www.kaggle.com/datasets/middlehigh/north-american-hurricanes-from-2000) dataset to develop a model of the factors driving damage costs. Of these factors:

**I predict that time has an effect on damage costs (increasing over time).**

- H0: Time has no effect on damage costs.
- HA: Time has an effect on damage costs.

## Explore and Clean Data

### Import packages
```{r}
#| code-summary: Expand Code
# Load required packages
library(tidyverse)
library(here)
library(janitor)
library(patchwork)
library(kableExtra)
library(webshot2)
```

### Import data
```{r}
#| code-summary: Expand Code
# Remove scientific notation
options(scipen=999)

# Import hurricane data
hurricane_data <- read_csv(here("data", "Hurricane Data.csv")) %>%
  clean_names()
```

### Inspect NA values for the response variable

How are the NA values for `damage_usd` distributed across hurricane categories?
```{r}
#| code-summary: Expand Code
# Total storms by category
total_storms <- hurricane_data %>%
  group_by(category) %>%
  summarise(count = n()) %>%
  rename(Category = category,
         Count = count)

print(paste("There are", sum(is.na(hurricane_data$damage_usd)), "NA values associated with damage cost."))

# Check NA values for damage (by category)
na_storms <- hurricane_data %>%
  filter(is.na(damage_usd))%>%
  group_by(category) %>%
  summarise(count = n()) %>%
  rename(Category = category,
         Count = count)

na_table <- left_join(total_storms, na_storms, by = "Category") %>%
  rename("Total Storms" = Count.x,
         "NA Damage" = Count.y) %>%
  arrange(factor(Category, levels = c('TS', 'Category 1', 'Category 2', 'Category 3', 'Category 4', 'Category 5'))) %>%
  kbl() %>%
  kable_styling()

na_table
```
Since NA values are somewhat proportionally distributed, it appears that removing them will not have an impact on the analysis. Additionally, I will scale `damage_usd` to be represented by "Millions of USD", saved as `damage_mil`.

```{r}
#| code-summary: Expand Code
# Remove rows with NA values for damage costs
hurricane_data_cleaned <- hurricane_data[!is.na(hurricane_data$damage_usd),]

# Find total number of areas affected
hurricane_data_cleaned <- hurricane_data_cleaned %>%
  separate_longer_delim(affected_areas, ",") %>%
  group_by(year, name, category, rain_inch, highest_wind_speed, damage_usd, fatalities) %>%
  summarise(total_areas = n()) %>%
  # Scale down damage costs
  mutate(damage_mil = damage_usd/1000000,
         time = year - 2000)
```

# Visualize
```{r}
#| code-summary: Expand Code
# Cost of damage by category
ggplot(hurricane_data_cleaned, aes(x = factor(category, levels = c("TS", "Category 1", "Category 2", "Category 3", "Category 4", "Category 5"), labels = c("TS", "1", "2", "3", "4", "5")), y = damage_mil)) +
  geom_boxplot() +
  labs(x = "Category",
       y = "Damage (Millions of USD)",
       title = "Cost of Storm Damage by Category 2000-2023") +
  theme_minimal()

ggsave(here("graphs", "damage_category_boxplot.jpg"))

mean_median <- hurricane_data_cleaned %>%
  group_by(category) %>%
  summarise(median = median(damage_usd/1000000),
            mean = mean(damage_usd/1000000)) %>%
  arrange(factor(category, levels = c('TS', 'Category 1', 'Category 2', 'Category 3', 'Category 4', 'Category 5'))) %>%
  rename(Category = category,
         "Median Damage Costs" = median,
         "Mean Damage Costs" = mean) %>%
  kbl() %>%
  kable_styling()

mean_median
```

# Visualize

Since category 5 storms include wind speeds of 157+ mph, it is possible that increasing `damage_mil` are associated with an increasing number of category 5 storms. The open-ended category will amass more storms as wind speed (`highest_wind_speed`) is expected to increase over time.

```{r}
#| code-summary: Expand Code
#| message: false
# Wind speed over time
ggplot(hurricane_data_cleaned, aes(x = year, y = highest_wind_speed)) +
  geom_point() +
  labs(x = "Year",
       y = "Max Wind Speed (mph)",
       title = "Max Wind Speed 2000-2023") +
  geom_hline(yintercept =157,
             linetype = "dashed",
             color = "red") +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1) +
  theme_minimal()
```

You can see how the time period 2000-2014 had 2 category 5 storms while the time period 2015-2023 has 8. 

Since rainfall is also expected to increase with climate change, we might want to include `rain_inch` it in our model.

```{r}
#| code-summary: Expand Code
#| message: false
# Rainfall over time
ggplot(hurricane_data_cleaned, aes(x = year, y = rain_inch)) +
  geom_point() +
  labs(x = "Year",
       y = "Rainfall (in)",
       title = "Rainfall 2000-2023") +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1) +
  theme_minimal()
```

The last factor that may be playing a role in `damage_mil` is the number of places affected (`total_areas`), calculated by counting how many "areas" are listed in the `areas_affected` column from the original dataset. As storms increase in intensity, I would expect that they will impact more places over time; however, we see that this may not be the case. We will still explore a potential relationship between `total_areas` and `damage costs`.

```{r}
#| code-summary: Expand Code
#| message: false
# Number of places over time
ggplot(hurricane_data_cleaned, aes(x = year, y = total_areas)) +
  geom_point() +
  labs(x = "Year",
       y = "Number of Places Affected",
       title = "Number of Places Affected 2000-2023") +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1) +
  theme_minimal()
```

Max wind speed, rainfall, number of areas affected, and time are interesting factors that are likely driving damage costs. Let's visualize their individual relationships with damage costs.

```{r}
#| code-summary: Expand Code
#| message: false
# Rainfall over time
wind_damage <- ggplot(hurricane_data_cleaned, aes(x = highest_wind_speed, y = damage_mil)) +
  geom_point() +
  labs(x = "Max Wind Speed (mph)",
       y = "Damage Costs",
       title = "Damage Costs vs. Wind") +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1) +
  theme_minimal()

rain_damage <- ggplot(hurricane_data_cleaned, aes(x = rain_inch, y = damage_mil)) +
  geom_point() +
  labs(x = "Rainfall (in)",
       y = "Damage Costs",
       title = "Damage Costs vs. Rainfall") +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1) +
  theme_minimal()

# Number of places vs. damage costs
areas_damage <- ggplot(hurricane_data_cleaned, aes(x = total_areas, y = log(damage_mil))) +
  geom_point() +
  labs(x = "Number of Places Affected",
       y = "Damage Costs",
       title = "Damage Costs vs. Places") +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1) +
  theme_minimal()

# Damage costs over time
time_damage <- ggplot(hurricane_data_cleaned, aes(x = time, y = log(damage_mil))) +
  geom_point() +
  labs(x = "Time (Years since 2000)",
       y = "Damage Costs",
       title = "Damage Costs vs. Time") +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1) +
  theme_minimal()

wind_damage + rain_damage + areas_damage + time_damage
```

# Analysis

We will be testing my hypothesis by normal approximation of the regression coefficient for `time`. In addition to `time`, we must include all relevant predictor variables in our model to reduce the risk of omitted variable bias:

- `highest_wind_speed`: highest wind speed achieved by the storm (mph)
- `rain_inch`: rain that fell (inches)
- `total_areas`: a count of the number of places listed in `areas affected`
- `time`: years since 2000

## Multiple Regression Model
```{r}
#| code-summary: Expand Code
# Create the model
damage_model <- lm(damage_mil ~ highest_wind_speed + rain_inch + total_areas + time, data = hurricane_data_cleaned)
summary(damage_model)
```

### Extract p-values
```{r}
#| code-summary: Expand Code
# p-values
beta1_p <- summary(damage_model)$coefficients[2,4]
beta2_p <- summary(damage_model)$coefficients[3,4]
beta3_p <- summary(damage_model)$coefficients[4,4]
beta4_p <- summary(damage_model)$coefficients[5,4]

beta <- c(beta1_p, beta2_p, beta3_p, beta4_p)

# Print p-values
for (i in seq_along(beta)) {
  print(paste0("The p-value for Beta ", i, " is ", beta[i], "."))
}
```

Now, let's visualize these relationships. Using mean rainfall and mean number of places affected, we can look at the relationship between time and damage costs by wind speed. 
```{r}
#| code-summary: Expand Code
# Update model
predictions <- expand_grid(
  highest_wind_speed = c(74, 96, 111, 130, 157),
  rain_inch = mean(hurricane_data_cleaned$rain_inch),
  total_areas = mean(hurricane_data_cleaned$total_areas),
  time = seq(0, 23, length.out = 100)) %>%
  mutate(damage_predicted = predict(damage_model, 
                                      newdata = .,
                                      type = "response"))

#Visualize
ggplot(predictions, aes(time, damage_predicted, color = factor(highest_wind_speed))) +
  geom_line() +
  scale_color_brewer(palette = "Reds",
                     name ="Wind Speed (mph)",
                   labels = c("74 (Cat 1)", "96 (Cat 2)", "111 (Cat 3)", "130 (Cat 4)", "157 (Cat 5)")) +
   labs(x = "Years Since 2000",
       y = "Predicted Damage Costs (Millions of USD)",
       title = "Predicted Damage Costs over Time by Wind Speed") +
  theme_bw()
```

```{r}
#| code-summary: Expand Code
# Update model
predictions3 <- expand_grid(
  highest_wind_speed = mean(hurricane_data_cleaned$highest_wind_speed),
  rain_inch = c(quantile(hurricane_data_cleaned$rain_inch)[2],
                   quantile(hurricane_data_cleaned$rain_inch)[3],
                   quantile(hurricane_data_cleaned$rain_inch)[4]),
  total_areas = mean(hurricane_data_cleaned$total_areas),
  time = seq(0, 23, length.out = 100)) %>%
  mutate(damage_predicted = predict(damage_model, 
                                      newdata = .,
                                      type = "response"))

# Visualize
ggplot(predictions3, aes(time, damage_predicted, color = factor(rain_inch))) +
  geom_line() +
  scale_color_brewer(palette = "Reds",
                     name ="Rainfall (in)") +
   labs(x = "Years Since 2000",
       y = "Predicted Damage Costs (Millions of USD)",
       title = "Predicted Damage Costs over Time by Rainfall Quantiles") +
  theme_bw()
```

```{r}
#| code-summary: Expand Code
# Update model
predictions4 <- expand_grid(
  highest_wind_speed = mean(hurricane_data_cleaned$highest_wind_speed),
  rain_inch = mean(hurricane_data_cleaned$rain_inch),
  total_areas = seq(min(hurricane_data_cleaned$total_areas),
                   max(hurricane_data_cleaned$total_areas),
                   by = 5),
  time = seq(0, 23, length.out = 100)) %>%
  mutate(damage_predicted = predict(damage_model, 
                                      newdata = .,
                                      type = "response"))

# Visualize
ggplot(predictions4, aes(time, damage_predicted, color = factor(total_areas))) +
  geom_line() +
  scale_color_brewer(palette = "Reds",
                      name ="Number of Places") +
   labs(x = "Years Since 2000",
       y = "Predicted Damage Costs (Millions of USD)",
       title = "Predicted Damage Costs over Time by Places") +
  theme_bw()
```

## Conclusions

I found that the p-values for Beta 1 (`hightest_wind_speed`) and Beta 2 (`rain_inch`) were less than $alpha=0.05$.

Beta 3 (`total_areas`) and Beta 4 (`time`) were greater than $alpha=0.05$. 

While we cannot confirm that damage costs are affected by wind speed or rainfall, we can rule out that their influence is due to random chance. 

As for time, we failed to reject the null hypothesis: 

- H0: Time has no effect on storm damage costs.

So, to conclude, I could not rule out that the effect of `time` on `damage_mil` could be due to random chance. 

There are a few factors are not considered in these models, such as the effects of inflation. There is likely an interaction between time and inflation, 

Additional predictor variables that would be of interest:

- storm surge
- water temperature

## Next steps

Given more time, I would have been interested in compiling a better dataset. The dataset used in this analysis isn't the best (no original source or thorough descriptions of variables i.e. rainfall?).

https://www.ncei.noaa.gov/access/billions/

# References

https://stock.adobe.com/search/images?k=hurricane
